# Reproducible Research: Peer Assessment 1
This assignment uses a sample dataset from a personal activity monitoring device that was provided through the R. Peng's github repository:
https://github.com/rdpeng/RepData_PeerAssessment1

I forked this repository and then used the provided template file to create this summary writeup. First the knitr and lattice packages are required and must be loaded.

```{r results='hide'}
library(knitr)
library(lattice)
```
## Loading and preprocessing the data
The dataset is a comma delimited file found in "activity.zip". This archive file is unzipped and the contents are read into a data frame with column classes specified.
```{r}
dzipfile <- "activity.zip"
unzip(dzipfile)
dfile <- "activity.csv"
df <- read.csv(dfile, header=TRUE, colClasses=c("numeric", "character", "integer"),
               na.strings="NA")
head(df)
tail(df)
summary(df)
```
From the summary of the data frame, we see that there are `r sum(is.na(df$steps))` missing values out of `r nrow(df)` rows of data. We will deal with missing values in a later section below.


## What is mean total number of steps taken per day?
The data give the number of steps recorded in each five minute interval of each date. There are a number of missing values, and some exploration shows that all the missing values are limited to 8 particular days which have no recorded number of steps at any time of those 8 days.

First we are interested in the total number of steps per day. We calculate the sum by day, and plot a histogram of the number of steps per day. Then, we calculate the mean of the total number of steps per day, with the missing values removed. The median is also calculated for comparison to a later calculation. 
```{r}
stepsPerDay <- with(df, tapply(steps, date, sum))

hist(stepsPerDay, breaks=8, main="Mean number of steps per day",
        xlab="Number of steps")
rug(stepsPerDay)

summary(stepsPerDay)
mean(stepsPerDay, na.rm=TRUE)
median(stepsPerDay, na.rm=TRUE)
```

## What is the average daily activity pattern?
Next, we want to see if there is a pattern to the daily activity, so we will average the number of steps for each 5-minute interval across all of the days, and plot this average number of steps vs. time interval. We find the time interval of the maximum number of steps on average, and show this on the figure as well.
```{r}
avgDailyPattern <- with(df, tapply(steps, interval, mean, na.rm=TRUE))
summary(avgDailyPattern)
x5minInterval <- df$interval[1:288]        
maxStepsInterval <- x5minInterval[which(avgDailyPattern==max(avgDailyPattern))]  
maxStepsInterval
max(avgDailyPattern)

plot(x5minInterval, avgDailyPattern, type="l",
     xlab="time interval (min)", ylab="daily average # steps",
     main="average daily pattern")   
text(1500, 200, paste("maximum =", 
                      sprintf("%4.1f", max(avgDailyPattern)), "at time =",
                      maxStepsInterval))
```
                      
## Inputing missing values
As noted earlier, there are missing values for 8 of the days. These 8 days are entirely empty, and will be replaced with the average number of steps by 5-mintue interval, which we calculated above. The histogram of this modified dataset with missing values replaced is then replotted.
```{r}
numMissing <- sum(is.na(df$steps))
numMissing
dfModified <- df

for (i in 1:length(dfModified$steps)){
  if(is.na(dfModified$steps[i])){
    dfModified$steps[i] <- avgDailyPattern[which(x5minInterval==dfModified$interval[i])]
  }
}
stepsPerDayMod <- with(dfModified, tapply(steps, date, sum))

hist(stepsPerDayMod, breaks=8, main="Mean number of steps per day, missing data input", xlab="Number of steps")
rug(stepsPerDayMod)
```
The median and mean of the modified dataset are calculated and compared against the median and mean of the original dataset. Notice that the mean values are unchanged by the modification for missing values, but the median has shifted to a slightly higher value.
```{r}
summary(stepsPerDayMod)
mean(stepsPerDayMod)
median(stepsPerDayMod)

summary(stepsPerDay)
mean(stepsPerDay, na.rm=TRUE)
median(stepsPerDay, na.rm=TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?
Finally we are interested to learn if there are differences in activity patterns on weekdays compared to weekends. So, we average the number of steps in 5-minute intervals across days by type: weekday or weekend. Then, we plot the average number of steps by interval in a panel plot to visually compare the weekday values to weekend values. We can see that on weekdays, the number of steps begins to increase at an earlier time, indicating that this person rises earlier on weekdays. Also, the number of steps decreases at a later time on weekends, indicating that the person goes to sleep at a later hour on weekends. 
```{r}
dfModified$type <- "weekday"
for (i in 1:length(dfModified$steps)){
  if(weekdays(as.Date(dfModified$date[i], "%Y-%m-%d"))=="Saturday" |
     weekdays(as.Date(dfModified$date[i], "%Y-%m-%d"))=="Sunday"){
    dfModified$type[i] <- "weekend"
  }
}
dfModified$type <- as.factor(dfModified$type)
summary(dfModified$type)

avgWDPattern <- with(dfModified[dfModified$type=="weekday",], tapply(steps, interval, mean, na.rm=TRUE))
avgWEPattern <- with(dfModified[dfModified$type=="weekend",], tapply(steps, interval, mean, na.rm=TRUE))

x5minInterval2 <- dfModified$interval[1:288]

avgsteps <- c(avgWDPattern, avgWEPattern)
avgInt <- c(x5minInterval2, x5minInterval2)
avgType <- ""
avgType[1:288] <- "weekday"
avgType[289:576] <- "weekend"
avgType <- as.factor(avgType)
dfnew <- data.frame(avgInt, avgsteps, avgType)


xyplot(avgsteps ~ avgInt|avgType, data = dfnew, layout=c(1,2),
       type="l", main="Weekend vs. Weekday Activity Patterns", 
       xlab="Interval", ylab="Number of steps")
```

